---
title: "Keynes' Consumption-Income Theory"
author: 
  - name: "Tomasz Przechlewski"
    email:  t.plata-przechlewski@psw.kwidzyn.edu.pl
    affiliation: "Powiślańska Szkoła Wyższa (Kwidzyn/Poland)"
description: (c) Tomasz Przechlewski / CC-BY license 
date: "2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("tidyverse") ## data manipulation
library('knitr')     ## decent tables
library("lmtest")    ##
library("car")       ## D-W test
library('eurostat')
```


## The problem 

Keynes observed that: *As income increases, consumption also 
increases, but by a smaller amount*:

$$C = a + b Y$$

where: 

$C$ -- consumption;

$a$ -- autonomous consumption (consumption when income is zero);

$b$ -- marginal propensity to consume (MPC), where $0< b <1$

$Y$ -- Income

## The data

From Eurostat's `nama_10_gdp` table we extract:
`P3` (*Final consumption expenditure*) 
and 'B1GQ' (*Gross domestic product at market prices*) 
for EU members in 2023 using R's `eurostat` package.

##

```{r, echo=FALSE, warning=FALSE, message=FALSE}
EU.members <- c("AT", "BE", "BG", "CY", "CZ", "DE", "DK", "EE", 
                "EL", "ES", "FI", "FR", "HR",
                "HU", "IE", "IT", "LT", "LU", "LV", "MT", "NL", 
                "PL", "PT", "RO", "SE", "SI", "SK")

## Income quintile share ratio S80/S20 by NUTS 2 region
## EU only

## GDP
## na_item
## P3 == final consumption expenditure
## B1GQ == GDP
gdp <- get_eurostat('nama_10_gdp') %>%
  ## Current prices EUR
  filter(geo %in% EU.members) |>
  filter (unit == 'CP_MEUR') |>
  filter (na_item == 'P3' | na_item == 'B1GQ') %>%
  mutate (year = as.numeric( substr(TIME_PERIOD, 1, 4))) %>%
  select (year, geo, na_item, values) |>
  pivot_wider(##id_cols = c('year', 'geo'),
    names_from = 'na_item', values_from = 'values') |>
  filter (year == 2023) |>
  select (geo, c=P3, gdp=B1GQ)
```

## The chart

```{r, echo=FALSE, warning=FALSE, message=FALSE}
p1 <- gdp |> ggplot(mapping=aes(x=gdp, y=c)) +
  geom_point(size=1) +
  geom_smooth(method='lm', fill=NA) 
p1
```


## Testing hypothesis on MPC

Estimating with **regression method**:

```{r}
m0 <- lm(c ~ gdp, data=gdp)
summary(m0)
```

Thus the estimated consumption function for `r nrow(gdp)` EU members
is:

$$ \hat y = 10864.72 + 0.76 \mathrm{GDP}$$

Keynes' hypothesis was confirmed!